---
title: "Analyzing Baseball Data with R - Chapter 1d"
format: 
  html:
    self-contained: true
---

```{r}
#| message: false
#| include: false

# libraries 
library(tidyverse)
library(Lahman)
library(patchwork)

# theme 
gg_theme = theme_minimal()+
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(face = "italic"),
    axis.title = element_text(face = "italic"),
    legend.position = "bottom"
  )
```


## Lahman Database Questions

### Home Runs and Strikeouts 
*What is the averaged number of home runs per game recorded in each decade? Does the rate of strikeouts show any correlation with the rate of home runs?*

First, we need to get the data for home runs and strike outs per game.  
```{r}
#| message: false
#| include: true

# get Lahman team data 
teams = Lahman::Teams

# get home runs by decade 
home_run_by_decade = teams %>% 
  mutate(decade = paste(substr(yearID, 1, 3), 0, sep = "")) %>% 
  group_by(decade) %>% 
  summarize(total_hr = sum(HR), total_g = (sum(G)/2), total_k = sum(SO, na.rm = T), hrpg = total_hr/total_g, kpg = total_k/total_g) %>% 
  ungroup()
```

Now we can graph this table to look at the three questions of interest. First, what is the average number of home runs per game for each decade. 

```{r}
#| message: false
#| code-fold: true

# graph of home runs by decade 
ggplot(home_run_by_decade, aes(decade, hrpg))+
  geom_point()+
  geom_line(aes(group = 1))+
  theme_minimal()+
  labs(
    title = "Home Runs Per Game by Decade",
    x = "Decade",
    y = "Home Runs Per Game"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 16)
  )

```
Home runs have been increasing over the history of Major League Baseball. In the first 40 years of the game, home runs happened rarely - 0.5 per game. In the early 1900s, home runs began to jump crossing over one per game in the 1930s. The rate of home runs has continued to increase, jumping to over 1.75 in the 1950s and over 2 per game in the 1990s. 

```{r}
#| message: false
#| code-fold: true

# strike outs per game by decade 
ggplot(home_run_by_decade, aes(decade, kpg)) + 
  geom_point() + 
  geom_line(group = 1) +
  theme_minimal() +
  labs(
    title = "Strike Outs Per Game",
    x = "Decade", 
    y = "Strikeouts Per Game"
  ) + 
  theme(
    plot.title = element_text(face = "bold", size = 20)
  )
```
Strike outs have increased over time in a slightly different manner. For the first 70 years, the rate of strikeouts was relatively stable. After jumping to around 7.5 per game in the 1880s, strikeouts stabilized between 6 and 7 per game. From 1920 onward, each decade has featured more strikeouts than the last with one exception - in 1969 MLB moved the mound back a foot, leading to less strikeouts in the 1970s. 

Plotted next to one another, you can see the clear correlation between the two over the decades. 

```{r}
#| message: false
#| code-fold: true

# both mapped together 
plot1 = ggplot(home_run_by_decade, aes(as.numeric(decade), hrpg))+
  geom_point(color = "darkblue")+
  geom_line(aes(group = 1), color = "darkblue")+
  theme_minimal()+
  labs(
    title = "Relationship Between Home Runs and Strikeouts by Decade",
    subtitle = "Home Runs per Game",
    x = NULL, 
    y = NULL
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(face = "italic")
  )

plot2 = ggplot(home_run_by_decade, aes(as.numeric(decade), kpg)) + 
  geom_point(color = "darkred") + 
  geom_line(group = 1, color = "darkred") +
  theme_minimal() +
  labs(
    subtitle = "Strikeouts Per Game",
    x = NULL, 
    y = NULL
  ) + 
  theme(
    plot.subtitle = element_text(face = "italic")
  )

plot1 + plot2
```

When you look at the percent increase across the decades, again, it is easy to spot the correlation between these two game events. From 1970 onward, the two have mirrored one another very closely. 
```{r}
#| warning: false
#| message: false
#| code-fold: true

# looking at percent change 
home_run_by_decade %>% 
  mutate(hr_delta = (hrpg - lag(hrpg))/lag(hrpg)*100,
         k_delta = (kpg - lag(kpg))/lag(kpg)*100) %>% 
  select(decade, hr_delta, k_delta) %>% 
  mutate(decade = as.numeric(decade)) %>% 
  pivot_longer(cols = hr_delta:k_delta, names_to = "stat", values_to = "value") %>% 
  ggplot(aes(x = decade, y = value, color = stat))+
  scale_color_manual(values = c("darkblue","darkred"), labels = c("Home Runs", "Strikeouts"))+
  geom_point()+
  geom_line()+
  theme_minimal()+
  labs(
    title = "Percent Change in HRs and Ks by Decade",
    x = "Decade",
    y = "Percent Change",
    color = NULL
  )+
  theme(
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "italic"),
    legend.position = "bottom"
  )
```

### The Implementation of the Designated Hitter
*What effect has the introduction of the Designated Hitter (DH) in the American League had in the difference in run scoring between the American and National Leagues?*

Let's explore this a couple of ways. First, lets look simply at the league wide run scoring. We can find this data in the `Teams` table from the `Lahman` package.

```{r}
Lahman::Teams %>%
  filter(lgID %in% c("NL", "AL")) %>% 
  group_by(lgID, yearID) %>% 
  summarize(r = sum(R)) %>% 
  ungroup() %>% 
  filter(yearID > 1959) %>% 
  ggplot(aes(x = yearID, y = r, color = lgID))+
  geom_line()+
  geom_vline(xintercept = 1973, color=  "red")+
  geom_vline(xintercept = 1981, linetype = "dashed", alpha = 0.2)+
  geom_vline(xintercept = 1994, linetype = "dashed", alpha = 0.2)+
  geom_vline(xintercept = 2020, linetype = "dotted", alpha = 0.2)+
  annotate(geom = "text", x = 1985, y = 12100, label = "1981 Strike", alpha = 0.2)+
  annotate(geom = "text", x = 1998, y = 13000, label = "1994 Strike", alpha = 0.2)+
  scale_color_manual(values = c("darkred", "darkblue"))+
  labs(
    x = "Year",
    y = "Total Runs Scored",
    title = "Runs Scored by League",
    subtitle = "1960 - Today",
    color = "League"
  )+
  gg_theme
```


## Appendix: 
Strikouts and home runs plotted on the same axis. 
```{r}
#| message: false
#| code-fold: true

home_run_by_decade %>% 
  pivot_longer(cols = hrpg:kpg, names_to = "stat", values_to = "spg") %>% 
  mutate(decade = as.numeric(decade)) %>%
  ggplot(aes(decade, spg, color = stat))+
  geom_point()+
  geom_line()+
  scale_color_manual(values = c("darkblue","darkred"), labels = c("Home Runs", "Strikouts"))+
  theme_minimal()+
  labs(
    title = "Relationship Between Home Runs and Strikeouts by Decade",
    x = "Decade",
    y = "# per Game",
    color = NULL
  )+
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```
